---
title: "Analysing Longitundinal Clinical Trials Data"
output: 
  html_notebook:
    toc: yes
  html_document:
    toc: yes
    df_print: paged
---

### Assessing the difference in moods from a drug across 6 different time points

# Using R

```{r}
# Load required libraries
library(lme4)  # for fitting linear mixed models
library(psych)
library(tidyverse)
library(ggplot2)
library(dplyr)


set.seed(42)  # For reproducibility

```

## Simulate data
```{r}
n_subjects <- 150
n_timepoints <- 6

# Create DataFrame with subjects, time points, and drug conditions
data <- data.frame(
  Subject = rep(1:n_subjects, each = n_timepoints),
  Time = rep(1:n_timepoints, times = n_subjects),
  Drug = sample(c(0, 1), n_subjects * n_timepoints, replace = TRUE)  # 0 = placebo, 1 = drug
)

# Simulate mood data
# Simulate mood data
data$Mood <- 6 + 0.3 * data$Time + 2.5 * data$Drug + 0.5 * data$Time * data$Drug + rnorm(n_subjects * n_timepoints)

# View first few rows of the dataset
head(data)
```

```{r}
describe(data)
```


```{r}
# Center and scale the 'Time' variable
data$Time_scaled <- scale(data$Time, center = TRUE, scale = TRUE)
```

The time variable should be centered and scaled to help with interpretation!


## 2. Fit Linear Mixed-Effects Models
```{r}
# Fit a mixed-effects model with random intercepts for Subject
model_1 <- lmer(Mood ~ Time_scaled + Drug + (1 | Subject), data = data)

# View model summary
summary(model_1)
```

This model tests mood as a function of Time and Drug.

Interpretation: Random intercepts allow each subject to have their own baseline mood level, 
while the fixed effects (Time and Drug) describe overall trends across all subjects.

Random effects:
The variance close to 0 indicates there's almost no variability in baseline mood 
across individuals. The residual variance is the variability in mood scores that 
isn’t explained by the model’s fixed effects (Time and Drug).
Since the residual variance is low, this suggests that random effects aren’t 
contributing much to the model’s fit.

Fixed effects:
For every 1 SD increase in time, mood increases by 0.94 points (on average for the placebo group), 
suggesting that mood improves over time regardless of the drug.

If a subject is taking the drug (compared to placebo), their mood increases by 
4.23 units on average regardless of time. 
This is a large effect, implying that the drug has a substantial positive impact on mood.

Correlation of fixed effects:

This small correlation between Time_scaled and Drug indicates that there is 
almost no multicollinearity between these two variables, so their estimates 
are stable and independent of one another.


## 3. Fit a More Complex Model with Random Slopes
```{r}
# Fit a mixed-effects model with random intercepts and slopes for Time
model_2 <- lmer(Mood ~ Time_scaled * Drug + (Time_scaled | Subject), data = data)

# View model summary
summary(model_2)

```

This model tests mood as a function of Time, Drug, and their interaction (Time × Drug)

Interpretation: The fixed effects include the interaction between Time and Drug, 
allowing for a changing effect of the drug over time. The random effects
allows each subject to have their own baseline mood and a unique trajectory over time.

Random effects: 
The intercept variances indicates no variation in the baseline mood across subjects.
The time variance indicates some variation in how subjects’ mood changes 
over time, though the variation is very small. This means most subjects 
follow a similar trend in mood change over time.
The residual variance indicates the variability of mood not explained by the 
fixed or random effects.

Fixed effects:
Time_scaled: For every 1 standard deviation increase in time in the placebo group, the mood increases by 0.41 points

Drug: Being in the drug group increases mood by 4.23 points (compared to the placebo group), 
regardless of time. This is a large effect.

Time_scaled:Drug: This is the interaction effect between Time_scaled and Drug. 
It means that for those in the drug group, the effect of time on mood is 1.02 
points higher per SD increase in time compared to the placebo group. In other words, 
the mood improvement over time is much greater in the drug group than in the placebo group.

The high t-values indicate significance! 

Correlation of Fixed Effects:
Time_scaled and drug:  The negative correlation here suggests that, as the effect 
of time increases, the effect of the drug decreases. This might be related to 
the strong interaction effect, as the rate of mood improvement over time is much 
stronger in the drug group.

Singular fit: 
Since there is little variation in baseline mood across subjects, 
we might consider dropping the random intercept term from the model to simplify it
and prevent overfitting. 


```{r}
# Fit a mixed-effects model with random intercepts and slopes for Time removing
# the random intercept term for time

model_3 <- lmer(Mood ~ Time_scaled * Drug + (0+ Time_scaled | Subject), data = data)

# View model summary
summary(model_3)

```

Everything looks almost identical, but no more singularity warning!

## 4. Visualize!
```{r}
# Create a new data frame for predictions
new_data <- data %>%
  group_by(Subject) %>%
  summarise(Time_scaled = seq(min(Time_scaled), max(Time_scaled), length.out = 100),
            Drug = unique(Drug[1]))  # Use unique drug status for each subject

# Get predictions from the model
new_data$Predicted_Mood <- predict(model_3, newdata = new_data, re.form = ~(0 + Time_scaled | Subject))

# Plot the observed data and the fitted lines
ggplot(data, aes(x = Time_scaled, y = Mood, color = as.factor(Drug))) +
  geom_point(alpha = 0.3) +  # Points for observed data
  geom_line(data = new_data, aes(x = Time_scaled, y = Predicted_Mood, group = Subject), color = "gray") +  # Individual subject lines
  labs(title = "Mood Over Time by Drug Treatment",
       x = "Time (Scaled)",
       y = "Mood",
       color = "Drug") +
  theme_minimal()
```


## 5. Compare Models Using Likelihood Ratio Test
```{r}
# Compare models using ANOVA (likelihood ratio test)
anova(model_1, model_3)
```

Interpretation: AIC (Akaike Information Criterion): A lower AIC indicates a better 
model fit when comparing models. Model 3 has a lower AIC (2567.8) than Model 1 (2774.9), 
suggesting it fits the data better.

Actually, all measures indicate Model 3 is better than Model 1! 
In other words, it is critical to take into the account the interaction effect!

The very low p-value (< 0.001) strongly suggests that the interaction between 
time and drug treatment is important for explaining the variation in mood.

## 5. Check Model Diagnostics
```{r}
# Plot residuals to check assumptions
plot(residuals(model_3))

```

Interpretation: This function retrieves the residuals of the fitted model model_3. 
Residuals are the differences between the observed values (in your case, Mood) 
and the predicted values from the model. They indicate how well the model is fitting the data.

After visually inspecting  for any patterns or trends, we see random scatter around zero:
this indicates that the model is appropriately specified regarding the linearity and homoscedasticity assumptions.

```{r}
# Random effects plot
plot(ranef(model_3))

```
Interpretation: This function extracts the random effects from the fitted model model_3. 
Random effects are the components of the model that account for variations 
attributable to the grouping structure in the data (in this case, Subject). 
They capture individual differences in the intercepts or slopes based on the random 
effects specified in the model.

After looking for any notable patterns or clusters in the random effects estimates,
we see that the random effects are evenly distributed and vary  among subjects:
this supports the complexity of your model with random effects. 

